Evolution Reinforces Cooperation with the Emergence of Self-Recognition Mechanisms: an Empirical Study of the Moran process for the Iterated Prisoner's Dilemma

### 2017-11-01 main.tex
Correct detail about repetitions of Moran process.

Apart from the PSO Gambler strategy, which was trained using a particle swarm
optimisation algorithm, these strategies are trained with an evolutionary
algorithm that perturbs strategy parameters and optimizes the mean total score
against all other opponents \cite{affenzeller2009genetic}. They were trained to
win IPD tournaments by maximizing their mean total payoffs against a variety
of opponents. Variation is
introduced via mutation and crossover of parameters, and the best performing
strategies are carried to the next generation along with new variants. Similar
methods appear in the literature \cite{Ashlock2006}.
There has also been some work on strategies using an evolutionary in real time: 
in~\cite{Gaudesi2016}
an evolutionary algorithm is used to build a model of the opponent and attempt
to exploit any potential weakness. This sort of work is not considered here:
strategies are instead pre trained.

@@

TF1 has an initial handshake of CCD and cooperates if the opponent matches.
However if the opponent later defects, TF1 will respond in kind, so the
handshake is not permanent. Only one player (Prober 4 \cite{Prison1998}) manages to
achieve cooperation with TF1 after about 20 rounds of play. TF1 is functionally
very similar to a strategy known as ``Collective Strategy'', which has a
handshake of CD and cooperates with opponents that matched the handshake
until they defect, defecting thereafter if the opponent ever defects \cite{Li2009}.
This strategy was specifically designed for evolutionary processes.

@@

\begin{table}[!hbtp]
    \centering
        \scalebox{0.7}{\input{tex/memory_depth_count.tex}}
        \caption{Memory depth}
        \label{tbl:memory_depth_count}
\end{table}

@@

Each strategy pair is run for 1000 repetitions of the Moran process to fixation
with starting population distributions of $(1, N-1)$, $(N/2, N/2)$ and $(N-1 ,
1)$, for \(N\) from 2 through 14. The fixation probability is then empirically
computed for each combination of starting distribution and value of \(N\).  The
Axelrod library can carry out exact simulations of the Moran process. Since some
of the strategies have a high computational cost or are stochastic, samples are
taken from a large number of 200 turn match outcomes for the pairs of players
for use in computing fitnesses in the Moran process (i.e. a stochastic cache of
matches is used). This approach was verified to agree with unsampled
calculations to a high degree of accuracy in specific cases.  This is described
in Algorithms~\ref{alg:data_collection} and~\ref{alg:moran_process}.

@@

\begin{table}[!hbtp]
    \centering
    \input{tex/t_test_results_calculator_v_arrogant_q_learner.tex}
    \caption{\(p\) values resulting from a \(t\) test comparing the theoretic
        value with the simulated value of the Moran Process fixation
        probabilities for \textbf{stochastic} strategies: Calculator and
    Arrogant Q Learner.}
    \label{tab:comparison_stochastic}
\end{table}

@@

When $N=2$ the Moran process is effectively a measure of the distribution of 
relative
mean payoffs over all possible matches between two players. The strategy
that scores higher than the other more often will fixate more often. For \(N=2\)
the two cases of \(x_1\) and \(x_{N-1}\) coincide, but will be
considered separately for larger \(N\) in sections~\ref{sec:strong_invaders}
and~\ref{sec:strong_resistors}. The top 16 (10\%) strategies are
shown in
Table~\ref{tbl:summary_top_2} and figures showing the performance of all
strategies are available in the supplementary materials. 
The top five ranking strategies are:

@@

It can be seen that apart from CS, none of the strategies
of Table~\ref{tbl:summary_top_2} perform well for \(N\in\{3, 7, 14\}\). The new
top performing strategies are:

\begin{itemize}
    \item Grudger (which only performs well for \(N=3\)), starts by cooperating
        but will defect if at any point the opponent has defected.
    \item MEM2, an infinite memory strategy that switches between TFT, TF2T, and
        Defector \cite{Li2014}.
    \item TF3, the finite state machine trained specifically for Moran processes
        described in Section~\ref{sec:introduction}.
    \item Prober 4, a strategy which starts with a specific 20 move sequence of
        cooperations and defections \cite{Prison1998}. This initial sequence serves
        as approximate handshake.
    \item  PSO Gambler and Evolved Lookerup 2 2 2: are strategies that make use
        of a lookup table mapping the first 2 moves of the opponent as well as
        the last 2 moves of both players to an action. The PSO gambler is a
        stochastic version of the Lookerup which maps those states to probabilities of cooperating. The Lookerup was described in \cite{Knight2016}.
    \item The Evolved ANN strategies are neural networks that map a number of
	    attributes (first move, number of cooperations, last move, etc.) to
	    an action. Both of these have been trained using an evolutionary
	    algorithm.
    \item The Evolved FSM 16 is a 16 state finite state machine trained to
        perform well in tournaments.
\end{itemize}

Only one of the above strategies is stochastic although close inspection of the
source code of PSO Gambler shows that it makes stochastic decisions rarely, and
is functionally very similar to its deterministic cousin Evolved Looker Up.
The PSO Gambler Mem1 strategy is a memory one strategy that has been trained to
maximise its utility and does perform well.
Apart from TF3, the finite state machines trained specifically for
Moran processes do not appear in the top 5, while strategies trained for
tournaments do. This is due to the nature of invasion: most of the opponents
will initially be different strategies. The next section will consider the
converse situation.

@@

The payoff maximizing strategies typically will not defect before the opponent's
first defection, possibly because the training strategy collection contains some
strategies such as Grudger and Fool Me Once that retaliate harshly by defecting
for the remainder of the match if the opponent has more than a small number of
cumulative defections. Paradoxically it is advantageous to defect (as a signal)
in order to achieve mutual cooperation with opponents using the same strategy
but not with other opponents. Nevertheless an evolutionary process is able to
tunnel through the costs and risks associated with early defections to find more
optimal solutions, so it is not surprising in hindsight that handshaking
strategies emerge from the evolutionary training process.

A handshake requires at least one defection and there is
selective pressure to defect as few times as possible to achieve the
self-recognition mechanism. It is also unwise to defect on the first move as
some strategies additionally retaliate first round defections. So the
handshakes used by TF1, TF2, and CS are in some sense optimal.

@@

For all starting populations
\(i\in\{1, N/2, N-1\}\) the ranks of strategies are relatively stable across the
different values of \(N>2\) however for \(N=2\) there is a distinct difference.
This highlights that there is little that can be inferred about the evolutionary
performance of a strategy in a large population from its performance in a small
population. This is confirmed by the performance of the zero determinant strategies: while
some do rank relatively highly for \(N=2\) (ZD-extort-4 has rank 16) this rank
does not translate to larger populations.

@@

Training strategies to excel
at the Moran process leads to the evolution of cooperation, but only with like
individuals in the case of TF1 and TF2. This may have significant implications
for human social interactions such as the evolution of ingroup/outgroup mechanisms
and other sometimes costly rituals that reinforce group behavior.

@@

One final point to recognise: the large set of strategies used here does not in
itself constitute an authoritative set. Whilst it is not only large but very
diverse, the results (and rankings) presented might change given a different set
of strategies. A further piece of work could look at subgroups of strategies and
how they fair against other subgroups. Note that because of the open nature of
the work here (not only is the source code archived but so is the data) this and
any other further analysis is possible to carry out.

>>>>

Apart from the PSO Gambler strategy, which was trained using a particle swarm
optimisation algorithm, these strategies are trained with an evolutionary
algorithm that perturbs strategy parameters and optimizes the mean total score
against all other opponents \cite{affenzeller2009genetic}. They were trained to
win IPD tournaments by maximizing their mean total payoffs against a variety
of opponents. Variation is
introduced via mutation and crossover of parameters, and the best performing
strategies are carried to the next generation along with new variants. Similar
methods appear in the literature \cite{Ashlock2006}.
There has also been some work on strategies using an evolutionary algorithm in real time: 
in~\cite{Gaudesi2016}
an evolutionary algorithm is used to build a model of the opponent and attempt
to exploit any potential weakness. In this work all strategies resulting from
evolutionary algorithms are pre-trained.

@@

TF1 has an initial handshake of CCD and cooperates if the opponent matches.
However if the opponent later defects, TF1 will respond in kind, so the
handshake is not permanent. Only one player (Prober 4 \cite{Prison1998}) manages to
achieve cooperation with TF1 after about 20 rounds of play. TF1 is functionally
very similar to a strategy known as ``Collective Strategy'', which has a
handshake of CD and cooperates with opponents that matched the handshake
until they defect, defecting thereafter if the opponent ever defects \cite{Li2009}.
Collective Strategy was specifically designed for evolutionary processes.

@@

\begin{table}[!hbtp]
    \centering
        \scalebox{0.7}{\input{tex/memory_depth_count.tex}}
        \caption{Memory depths}
        \label{tbl:memory_depth_count}
\end{table}

@@

The Moran process is run for each strategy pair to fixation
with starting population distributions of $(1, N-1)$, $(N/2, N/2)$ and $(N-1 ,
1)$, for \(N\) from 2 through 14. The fixation probability is then empirically
computed for each combination of starting distribution and value of \(N\).  The
Axelrod library can carry out exact simulations of the Moran process. Since some
of the strategies have a high computational cost or are stochastic, samples are
taken from a large number of 200 turn match outcomes for the pairs of players
for use in computing fitnesses in the Moran process (i.e. a stochastic cache of
matches is used). This approach was verified to agree with unsampled
calculations to a high degree of accuracy in specific cases.  This is described
in Algorithms~\ref{alg:data_collection} and~\ref{alg:moran_process}.

@@

\begin{table}[!hbtp]
    \centering
    \input{tex/t_test_results_calculator_v_arrogant_q_learner.tex}
    \caption{\(p-\text{values}\) resulting from a \(t-\text{ test}\) comparing the theoretic
        value with the simulated value of the Moran Process fixation
        probabilities for \textbf{stochastic} strategies: Calculator and
    Arrogant Q Learner.}
    \label{tab:comparison_stochastic}
\end{table}

@@

When $N=2$ the fixation probabilities of the Moran process are effectively measures
of the distribution of  relative
mean payoffs over all possible matches between two players. The strategy
that scores higher than the other more often will fixate more often. For \(N=2\)
the two cases of \(x_1\) and \(x_{N-1}\) coincide, but will be
considered separately for larger \(N\) in sections~\ref{sec:strong_invaders}
and~\ref{sec:strong_resistors}. The top 16 (10\%) strategies are
shown in
Table~\ref{tbl:summary_top_2} and figures showing the performance of all
strategies are available in the supplementary materials. 
The top five ranking strategies are:

@@

It can be seen that apart from CS, none of the strategies for \(N=2\)
of Table~\ref{tbl:summary_top_2} perform well for \(N\in\{3, 7, 14\}\). The new
top performing strategies are:

\begin{itemize}
    \item Grudger (which only performs well for \(N=3\)), starts by cooperating
        but will defect if at any point the opponent has defected.
    \item MEM2, an infinite memory strategy that switches between TFT, TF2T, and
        Defector \cite{Li2014}.
    \item TF3, the finite state machine trained specifically for Moran processes
        described in Section~\ref{sec:introduction}.
    \item Prober 4, a strategy which starts with a specific 20 move sequence of
        cooperations and defections \cite{Prison1998}. This initial sequence serves
        as approximate handshake.
    \item  PSO Gambler and Evolved Lookerup 2 2 2: strategies that make use
        of a lookup table mapping the first 2 moves of the opponent as well as
        the last 2 moves of both players to an action.  PSO gambler is a
        stochastic version of  Lookerup which maps those states to probabilities
       of cooperating. Lookerup was described in \cite{Knight2016}.
    \item The Evolved ANN strategies are neural networks that map a number of
	    attributes (first move, number of cooperations, last move, etc.) to
	    an action. Both of these have been trained using an evolutionary
	    algorithm.
    \item Evolved FSM 16 is a 16 state finite state machine trained to
        perform well in tournaments.
\end{itemize}

Only one of the above strategies is stochastic although close inspection of the
source code of PSO Gambler shows that it makes stochastic decisions rarely, and
is functionally very similar to its deterministic cousin Evolved Looker Up.
PSO Gambler Mem1 is a stochastic memory one strategy that has been trained to
maximise its utility and does perform well.
Apart from TF3, the finite state machines trained specifically for
Moran processes do not appear in the top 5, while strategies trained for
tournaments do. This is due to the nature of invasion: most of the opponents
will initially be different strategies. The next section will consider the
converse situation.

@@

The payoff maximizing strategies typically will not defect before the opponent's
first defection, possibly because the training strategy collection contains some
strategies such as Grudger and Fool Me Once that retaliate harshly by defecting
for the remainder of the match if the opponent has more than a small number of
cumulative defections. Paradoxically for handshaking strategies
it is advantageous to defect (as a signal)
in order to achieve mutual cooperation with opponents using the same strategy
but not with other opponents. Nevertheless an evolutionary process is able to
tunnel through the costs and risks associated with early defections to find more
optimal solutions, so it is not surprising in hindsight that handshaking
strategies emerge from the evolutionary training process.

A handshake requires at least one defection and there is
selective pressure to defect as few times as possible to achieve the
self-recognition mechanism. It is also unwise to defect on the first move as
some strategies additionally retaliate in response to first round defections. So the
handshakes used by TF1, TF2, and CS are in some sense optimal.

@@

For all starting populations
\(i\in\{1, N/2, N-1\}\) the ranks of strategies are relatively stable across the
different values of \(N>2\) however for \(N=2\) there is a distinct difference.
This highlights that there is little that can be inferred about the evolutionary
performance of a strategy in a large population from its performance in a small
population. This is confirmed by the performance of the zero determinant strategies: while
some do rank relatively highly for \(N=2\) (ZD-Extort-4 has rank 16) this rank
does not translate to larger populations.

@@

Training strategies to excel
at the Moran process leads to the evolution of cooperation, but only with like
individuals in the case of TF1 and TF2. This may have significant implications
for various biological and social phenomena such as
human social interactions, particularly the evolution of ingroup/outgroup mechanisms
and other sometimes costly rituals that reinforce group behavior.

@@

One final point to recognise: the large set of strategies used here does not in
itself constitute an authoritative set. Whilst it is not only large but also very
diverse, the results (and rankings) presented might change given a different set
of strategies. A further piece of work could look at subgroups of strategies and
how they fair against other subgroups. Note that because of the open nature of
the work here (not only is the source code archived but so is the data) this and
any other further analysis is possible to carry out.
